### 2007-01-04 Documentation/part1.tex
Prior chapter added

double chiDCNorm(int exp, int rule, int np, double *x, double *errors) \\
\{ \\
  \hspace*{0.3cm} const EXP\_FAR = 0; const EXP\_NEAR = 1; \\
  \hspace*{0.3cm} int n\_bins = glbGetNumberOfBins(EXP\_FAR); \\
  \hspace*{0.3cm} double *true\_rates\_N = glbGetRuleRatePtr(EXP\_NEAR, 0); \\
  \hspace*{0.3cm} double *true\_rates\_F = glbGetRuleRatePtr(EXP\_FAR, 0); \\
  \hspace*{0.3cm} double signal\_fit\_rates\_N[n\_bins]; double signal\_fit\_rates\_F[n\_bins]; \\
  \hspace*{0.3cm} double signal\_norm\_N, signal\_norm\_F; \\
  \hspace*{0.3cm} int ew\_low, ew\_high, i; \\
  \hspace*{0.3cm} double emin, emax, fit\_rate;  double chi2 = 0.0; \\
  \\
   \hspace*{0.3cm} /* Request simulated energy interval and analysis energy window */ \\
   \hspace*{0.3cm} glbGetEminEmax(exp, \&emin, \&emax); \\
   \hspace*{0.3cm} glbGetEnergyWindowBins(exp, rule, \&ew\_low, \&ew\_high);

@@

  \hspace*{0.3cm} /* Loop over all bins in energy window */ \\
  \hspace*{0.3cm} signal\_norm\_F = 1.0 + x[0] + x[1]; \\
  \hspace*{0.3cm} signal\_norm\_N = 1.0 + x[0] + x[2]; \\
  \hspace*{0.3cm} for (i=ew\_low; i <= ew\_high; i++) \\
  \hspace*{0.3cm} \{ \\
  \hspace*{0.6cm}  /* Statistical part of chi2 for far detector */ \\
  \hspace*{0.6cm}  fit\_rate  = signal\_norm\_F * signal\_fit\_rates\_F[i]; \\
  \hspace*{0.6cm}  chi2 += likelihood(true\_rates\_F[i], fit\_rate, true\_rates\_F[i]); \\
   \\
  \hspace*{0.6cm}  /* Statistical part of chi2 for near detector  */ \\
  \hspace*{0.6cm}  fit\_rate  = signal\_norm\_N * signal\_fit\_rates\_N[i]; \\
  \hspace*{0.6cm}  chi2 += likelihood(true\_rates\_N[i], fit\_rate, true\_rates\_N[i]); \\
  \hspace*{0.3cm} \}

@@

\index{norm}{Projection!of manifold}
In principle, the precision of an individual parameter measurement including
 correlations in the $\chi^2$-approach can be obtained as the projection of 
the $n$-dimensional fit manifold onto the respective axis. Similarly, one can 
project the fit manifold onto a plane, such as the $\stheta$-$\deltacp$-plane,
 if one wants to show the allowed
region in this plane with all the other parameter correlations included. 
In practice, this projection is very difficult: a grid-based method would 
need $(N_{\mathrm{grid}})^n$ function calls of {\tt glbChiSys} to 
calculate the projection onto one dimension 
including the full $n$-parameter correlation, where 
$N_{\mathrm{grid}}$ is the number of points in each direction of the 
lattice. For example, taking only $N_{\mathrm{grid}}=20$ and 
$n=7$ (six oscillation parameters and matter density) would mean more 
than one billion function calls of {\tt glbChiSys}. One can easily imagine 
that these would be too many for any reasonable application.

\index{norm}{Minimizer}
The solution to this problem is using a $n$-dimensional, local 
minimizer for the projection instead of a grid-based method, where we will
illustrate this minimization process later. It turns out
that such a minimizer can include a full $6$-parameter correlation 
with of the order of $1\, 000$ function calls of {\tt glbChiSys}. For the 
minimization we use a derivative free method due to Powell in a 
modified~\cite{Brent:1973} version\footnote{Not needing derivatives is
highly desired since the event rate depends in a non-linear way on the
oscillation parameters, thus there is no easy analytical way
to obtain derivatives of the $\chi^2$ function.}.

@@

This approach also has one major disadvantage: There is \emph{no} such
thing as a global minimization algorithm or even an algorithm which guarantees
to find \emph{all} local minima of a function. In practice this means using 
 a local minimizer, one may end up in an unwanted local minimum and not in 
the investigated (possibly global) one or 
one may miss a local minimum which affects the results\footnote{NB -- 
Implementing a grid-based method which guarantee to find all local minima is
not straightforward either, to say the least.}. 
The only way out of this dilemma is to use some heuristic approach, \ie\
although one can not guarantee anything one can use schemes which work
in most cases and announce their failure loudly. In order to use such a
heuristic some (analytical or numerical) knowledge on the topology of 
the fit manifold is necessary. With this knowledge it is possible
to obtain an approximate position for each local minimum 
and thus to start the local minimizer 
close enough to the investigated minimum.
 Fortunately, this can be done quite straightforward in most cases, since 
the structure of the neutrino oscillation formulas does not cause very 
complicated topologies of the fit manifolds. Especially the simulation of 
reactor experiments and conventional beams or superbeams is rather simple 
with purely numerical
approaches. Neutrino factories have, especially for small values of
$\theta_{13}$, a much more complicated topology. In this case, results
of the many analytical discussions of this issue can be used. This means
 that one can implicitly use the analytical knowledge to obtain 
better predictions for the location of a minimum. One can easily imagine 
that the used methods then also depend
on the region of the parameter space. In this manual, we only use
examples with a neutrino factory, since some of these complications
can be illustrated there. Albeit the methods described here are neither
complete nor will they work everywhere in the parameter space. It is
in any case up to the user to make sure that the results are what he/she
thinks.

Some more words of warning with respect to results obtained by projecting
the $\chi^2$:
The results obtained this way are always only a \emph{upper} bound on the
value of the projected $\chi^2$ function, \ie\ an undiscovered minimum 
decreases the value of the the projected $\chi^2$ function. If the value
of the $\chi^2$ function in the missed minimum is larger than the
previously found ones it will not influence the value of the projected 
value. 
Thus, one can only run the danger to obtain a too optimistic solution 
if one does not find the other local minima appearing 
below the chosen confidence level. Thus, with this approach and proper usage,
 it should not possible to produce a too pessimistic solution. 
However, if one is not careful enough to locate
all local minima, one can easily produce too optimistic solutions.
This danger can be summarized as follows:
\begin{equation}
\mathrm{Too \, pessimistic \, result} \, < \, \underbrace{\mathrm{Real \, result}}_{\mathrm{Located \, by \, careful \, usage}} \, \le \, \mathrm{GLoBES \, result} \,  < \,  \mathrm{Too \, optimisitic \, result} \nonumber
\end{equation}

\index{norm}{External information!priors} 
\index{norm}{External information!input errors} 
\index{norm}{External information!starting values}
In many cases, the fit manifold is restricted by the knowledge 
from earlier experiments. For example, the knowledge on the solar 
parameters will in most cases be supplied by the solar neutrino experiments. 
If the external precision of a parameter is at the time of the 
measurement better than the one of the experiment itself, one usually will use
this external, better knowledge and impose a corresponding constraint
 on this parameter. This external knowledge may reduce the extension of the 
$n$-dimensional fit manifold in the respective direction. In the most 
extreme case, keeping all parameters but the measured one fixed 
in the analysis is equivalent to the assumption that all parameters 
are determined externally with infinitively high precisions. As this 
is a quite strong assumption, one should \emph{always} check the 
consequences of relaxing it and using realistic errors. Only if such
a test has demonstrated that the impact of the uncertainty on a 
given fit parameter is negligible it can be assumed as fixed safely. 
 The inclusion of external input in \GLOBES\ is done by the use of 
Gau\ss ian {\em priors}: We assume that an external measurement has 
determined the measured parameter to be at the central value 
(which we call {\em starting value}) with a $1 \sigma$ Gau\ss ian error 
(which we call {\em input error}). The explicit definition of these 
priors will be shown in the next section.

@@

\index{norm}{External information}
\index{norm}{External input| \see{External information}}
It is one of the strengths of the \GLOBES\ software to use external input 
in order to reduce the extension of the fit manifold with the knowledge 
from external
(earlier) measurements. The treatment of external input is done by the 
addition of Gau\ss ian so-called {\em priors} to the systematics-minimized 
$\chi^2$-function. For example, for the matter density, one obtains as the 
final projected $\chi^2_F$ after minimization over the matter density
scaling factor \index{norm}{Matter density!scaling factor} $\hat{\rho}$
\be
 \chi^2_F = \underset{\rho}{\mathrm{min}} \left( \chi^2(\hat{\rho}) +
 \frac{(\hat{\rho} - \hat{\rho}^0)^2}{\sigma_{\hat{\rho}}^2} \right).
\label{equ:priors}
\ee
This example is a very simple one, since in fact the
minimization is simultaneously performed over all priors and free oscillation 
parameters. In \equ{priors}, $\hat{\rho}^0$ is the {\em starting value} of the 
prior, and $\sigma_{\hat{\rho}}$ the $1 \sigma$ absolute (half width) 
{\em input error}. Thus, it is assumed that an external measurement has 
determined the matter density with a precision (input error) 
$\sigma_{\hat{\rho}}$ at the central value $\hat{\rho}^0$. Usually, 
the starting value is fixed at the best-fit value, and the input error 
to the $1 \sigma$ half width of the external measurement. For the matter 
density, $\hat{\rho}^0$  is usually set to $1.0$ corresponding to
the actual matter density profile such as given by the experiment
 definition file, and $\sigma_{\hat{\rho}}$ to the 
relative matter density uncertainty (\eg, $0.05$ for 5\% uncertainty).

In principle, one can set the priors for the matter density and all oscillation parameters. For example, if the disappearance channels of the experiment determine the leading oscillation parameters with unprecedented
precisions, one can omit the respective input errors. In \GLOBES , a
value of $0$ corresponds\footnote{to be precise, a value for the error
in between $-10^{-12}$ and $+10^{-12}$} to neglecting the prior. If, however,
 earlier external measurements provide better information, one can set their 
absolute precisions with the input errors. The starting values are usually 
chosen to be the best-fit values of this external experiments, such as
for the input from solar experiments. In some cases, it may be
necessary to adjust them, such as for artificial constraints to the
oscillation parameters. In other cases, minor modifications of the 
starting values can cause a faster convergence of the algorithm.
For example, for the investigation of the
opposite-sign solution, one can use the prior to constrain $\ldm$
in order to force the minimizer not to  fall into the (unwanted) 
true-sign solution. In this case, the starting value of $\ldm$ would 
be set to $\rho^0_{\ldm} = - \ldm$, and a $\sigma_{\ldm}$ of the order 
of $\ldm$ would be imposed. For the algorithm, it would 
then be rather difficult 
to converge into the unwanted true-sign solution. However, note that 
one should in this case check that the actually determined value 
for $\ldm$ after minimization is close enough to the guessed value $-\ldm$ 
in order to avoid significant artifical contributions of the priors to the 
final $\chi^2$. Alternatively one could re-run the minimizer with the
position of the previously found minimum as starting position but now
with switching off the constraint on $\ldm$.

\index{norm}{External information!input errors} 
\index{norm}{External information!starting values}
In order to set the starting values and input errors, two function have to
 be called {\em before the usage of any minimizer}:
\begin{function}
\GLBNS{SetCentralValues}
{\tt int glbSetCentralValues(const glb\_params in)} sets the starting values for all of the following minimizer calls to {\tt in}.
\end{function}
\begin{function}
\GLBNS{SetInputErrors}
{\tt int glbSetInputErrors(const glb\_params in)}
 sets the input errors for all of the following minimizer calls to {\tt in}.
 An input error of $0$ corresponds to not taking into account the
 respective prior.
\end{function}
Accordingly, there are functions to return the actually set starting values
and input errors:
\begin{function}
\GLBNS{GetCentralValues}
{\tt int glbGetCentralValues(glb\_params out)} writes the currently
set starting values to {\tt out}.
\end{function}
\begin{function}
\GLBNS{GetInputErrors}
{\tt int glbGetInputErrors(glb\_params out)}
 writes the currently set input errors to {\tt out}.
\end{function}
All functions take or return as many matter density parameters as there are initialized experiments. In addition, they return $-1$ if the operation
was not successful.

\index{norm}{External information!precision}
Eventually, a typical initialization of the external input with
$10\%$ external precisions for the solar 
parameters\footnote{In fact, accelerator-based long-baseline experiments 
are primarily sensitive to the product $\sin 2 \theta_{12} \cdot \sdm$, 
which means that these errors effectively add up to an error of this 
product of about $15\%$.}, 
and $5\%$ matter density uncertainties for all experiments looks like this:
\begin{quote}
{\tt
 glbDefineParams(input\_errors,theta12*0.1,0,0,0,sdm*0.1,0);\\  
 glbSetDensityParams(input\_errors,0.05,GLB\_ALL);\\
 glbSetCentralValues(true\_values);\\
 glbSetInputErrors(input\_errors);\\
}
\end{quote}
In this example, the starting values are set to 
the true (simulated) values. Remember that initially the 
matter density scaling factors are 
all $1.0$, which means that they do not need to be adjusted for the
starting values.

\index{norm}{Minimizer!priors}
\index{norm}{External information!priors}
\index{norm}{Advanced tricks}
Though the priors are an elegant way to treat external input, 
there are also some complications with priors. The following hints are for the
more advanced \GLOBES\ user:
\begin{enumerate}
\item
 The priors are only added once to the final $\chi^2$, no matter how
 many experiments there are simulated. This is already one reason
 (besides the minimization) why the sum of all projected $\chi^2$'s of 
 the individual experiments 
 cannot correspond to the $\chi^2$ of the combination of all experiments.
\item
 Priors are not used for parameters which are not minimized over, \ie, 
 kept fixed. This will be important together with arbitrary projections
 using \GLB{ChiNP}. A more subtile consequence is the comparison
 of fit manifold sections and projections for the solutions where
 the absolute minimum $\chi^2$ is larger than zero, \ie, degeneracies
 other than the true solution. In this case, the sections and
 projections are not  comparable if not corrected by the prior contributions, where the
 correction can be obtained as the $\chi^2$-difference at the minimum.
 For example, projecting the $\mathrm{sgn}(\ldm)$-degeneracy onto
 the $\theta_{13}$-$\deltacp$-plane and comparing it with the section
 (all other parameters fixed), the section region would in many cases be 
 larger than the projection region if the priors are not added to the
 section. At the true solution, this problem usually does not occur
 because the prior contributions are close to zero.    
\item
Currently, \GLOBES\ only supports Gau\ss ian priors for the individual
oscillation parameters. Especially for the solar parameters, this is only
 an approximation, since they are imposed on $\theta_{12}$ and not
 on $\sin 2 \theta_{12}$, $\sin 2 \theta_{12} \cdot \sdm$, or $\sin  
 \theta_{12}$. Later versions of  \GLOBES\ may include more alternatives.
 \end{enumerate}

@@

\begin{quote}
{\tt {\scriptsize
  /* Set starting values and input errors for all projections */  \\
  glbDefineParams(input\_errors,theta12*0.1,0,0,0,sdm*0.1,0);\\  
  glbSetDensityParams(input\_errors,0.05,GLB\_ALL);\\
  glbSetCentralValues(true\_values);\\
  glbSetInputErrors(input\_errors);\\
\\
  /* Define my own two-parameter projection for glbChiNP: Only deltacp is free! */ \\
  glbDefineProjection(th13\_projection,GLB\_FIXED,GLB\_FIXED,GLB\_FIXED,GLB\_FREE,GLB\_FIXED,GLB\_FIXED); \\
  glbSetProjection(th13\_projection); \\ 
\\
  /* Iteration over all values to be computed */ \\
  double x,res1,res2;     \\
  for(x=-4;x<-2.0+0.001;x=x+2.0/50) \\
  \{ \\
\hspace*{0.5cm}  /* Set fit value of stheta */ \\
\hspace*{0.5cm} glbSetOscParams(test\_values,asin(sqrt(pow(10,x)))/2,1); \\
    \\
\hspace*{0.5cm} /* Guess fit value for deltacp in order to safely find minimum */ \\
\hspace*{0.5cm} glbSetOscParams(test\_values,200.0/2*(x+4)*M\_PI/180,3); \\
 \\
\hspace*{0.5cm} /* Compute Chi2 for user-defined two-parameter correlation */ \\
\hspace*{0.5cm} res1=glbChiNP(test\_values,NULL,GLB\_ALL); \\
      \\
\hspace*{0.5cm} /* Compute Chi2 for full correlation: minimize over all but theta13 */ \\
\hspace*{0.5cm} res2=glbChiTheta13(test\_values,NULL,GLB\_ALL); \\
  \\
\hspace*{0.5cm} AddToOutput(x,res1,res2);\\
  \} \\

>>>>

double chiDCNorm(int exp, int rule, int np, double *x, double *errors) \\
\{ \\
  \hspace*{0.3cm} const EXP\_FAR = 0; const EXP\_NEAR = 1; \\
  \hspace*{0.3cm} int n\_bins = glbGetNumberOfBins(EXP\_FAR); \\
  \hspace*{0.3cm} double *true\_rates\_N = glbGetRuleRatePtr(EXP\_NEAR, 0); \\
  \hspace*{0.3cm} double *true\_rates\_F = glbGetRuleRatePtr(EXP\_FAR, 0); \\
  \hspace*{0.3cm} double signal\_fit\_rates\_N[n\_bins]; double signal\_fit\_rates\_F[n\_bins]; \\
  \hspace*{0.3cm} double signal\_norm\_N, signal\_norm\_F; \\
  \hspace*{0.3cm} int ew\_low, ew\_high, i; \\
  \hspace*{0.3cm} double emin, emax, fit\_rate;  double chi2 = 0.0;

   \hspace*{0.3cm} /* Request simulated energy interval and analysis energy window */ \\
   \hspace*{0.3cm} glbGetEminEmax(exp, \&emin, \&emax); \\
   \hspace*{0.3cm} glbGetEnergyWindowBins(exp, rule, \&ew\_low, \&ew\_high);

@@

  \hspace*{0.3cm} /* Loop over all bins in energy window */ \\
  \hspace*{0.3cm} signal\_norm\_F = 1.0 + x[0] + x[1]; \\
  \hspace*{0.3cm} signal\_norm\_N = 1.0 + x[0] + x[2]; \\
  \hspace*{0.3cm} for (i=ew\_low; i <= ew\_high; i++) \\
  \hspace*{0.3cm} \{ \\
  \hspace*{0.6cm}  /* Statistical part of chi2 for far detector */ \\
  \hspace*{0.6cm}  fit\_rate  = signal\_norm\_F * signal\_fit\_rates\_F[i]; \\
  \hspace*{0.6cm}  chi2 += likelihood(true\_rates\_F[i], fit\_rate, true\_rates\_F[i]); \\

  \hspace*{0.6cm}  /* Statistical part of chi2 for near detector  */ \\
  \hspace*{0.6cm}  fit\_rate  = signal\_norm\_N * signal\_fit\_rates\_N[i]; \\
  \hspace*{0.6cm}  chi2 += likelihood(true\_rates\_N[i], fit\_rate, true\_rates\_N[i]); \\
  \hspace*{0.3cm} \}

@@

\index{norm}{Projection!of manifold}
In principle, the precision of an individual parameter measurement including
 correlations in the $\chi^2$-approach can be obtained as the projection of 
the $n$-dimensional fit manifold onto the respective axis. Similarly, one can 
project the fit manifold onto a plane, such as the $\stheta$-$\deltacp$-plane,
 if one wants to show the allowed
region in this plane with all the other parameter correlations included. 
In practice, this projection (or marginalization) is very difficult: a grid-based method would 
need $(N_{\mathrm{grid}})^n$ function calls of {\tt glbChiSys} to 
calculate the projection onto one dimension 
including the full $n$-parameter correlation, where 
$N_{\mathrm{grid}}$ is the number of points in each direction of the 
lattice. For example, taking only $N_{\mathrm{grid}}=20$ and 
$n=7$ (six oscillation parameters and matter density) would mean more 
than one billion function calls of {\tt glbChiSys}. One can easily imagine 
that these would be too many for any reasonable application.

\index{norm}{Minimizer}
The solution to this problem is using a $n$-dimensional, local 
minimizer for the projection instead of a grid-based method, where we will
illustrate this minimization process later. It turns out
that such a minimizer can include a full $6$-parameter correlation 
with of the order of $1\, 000$ function calls of {\tt glbChiSys}. For the 
minimization we use a derivative free method due to Powell in a 
modified~\cite{Brent:1973} version\footnote{Not to need derivatives is
highly desired, since the event rate depends in a non-linear way on the
oscillation parameters. Thus, there is no easy analytical way
to obtain derivatives of the $\chi^2$ function.}.

@@

This approach also has one major disadvantage: There is \emph{no} such
thing as a global minimization algorithm or even an algorithm which guarantees
to find \emph{all} local minima of a function. In practice this means using 
 a local minimizer, one may end up in an unwanted local minimum and not in 
the investigated (possibly global) one or 
one may miss a local minimum which affects the results\footnote{NB -- 
Implementing a grid-based method which guarantees to find all local minima is
not straightforward either, to say the least.}. 
The only way out of this dilemma is to use some heuristic approach, \ie , 
one can use schemes which work
in most cases and announce their failure loudly. In order to use such a
heuristic some (analytical or numerical) knowledge on the topology of 
the fit manifold is necessary. With this knowledge it is possible
to obtain an approximate position for each local minimum 
and thus to start the local minimizer 
close enough to the investigated minimum.
 Fortunately, this can be done quite straightforward in most cases, since 
the structure of the neutrino oscillation formulas does not cause very 
complicated topologies of the fit manifolds. Especially the simulation of 
reactor experiments and conventional beams or superbeams is rather simple 
with purely numerical
approaches. Neutrino factories have, especially for small values of
$\theta_{13}$, a much more complicated topology. In this case, results
of the many analytical discussions of this issue can be used. This means
 that one can implicitly use the analytical knowledge to obtain 
better predictions for the location of a minimum. One can easily imagine 
that the used methods then also depend
on the region of the parameter space. In this manual, we mainly use
examples with a neutrino factory, since some of these complications
can be illustrated there. Note that the methods described here are neither
complete, nor will they work everywhere in the parameter space. It is
in any case up to the user to make sure that the results are what he/she
thinks.

Some more words of warning with respect to results obtained by projecting
the $\chi^2$:
The results obtained this way are always only a \emph{upper} bound on the
value of the projected $\chi^2$ function, \ie\ an undiscovered minimum 
decreases the value of the the projected $\chi^2$ function. If the value
of the $\chi^2$ function in the missed minimum is larger than the
previously found ones it will not influence the value of the projected 
value. 
Thus, one can only run the danger to obtain a too optimistic solution 
if one does not find the other local minima appearing 
below the chosen confidence level. With this approach and proper usage,
 it should not possible to produce a too pessimistic solution. 
However, if one is not careful enough to locate
all local minima, one can easily produce too optimistic solutions.
This danger can be summarized as follows:
\begin{equation}
\mathrm{Too \, pessimistic \, result} \, < \, \underbrace{\mathrm{Real \, result}}_{\mathrm{Located \, by \, careful \, usage}} \, \le \, \mathrm{GLoBES \, result} \,  < \,  \mathrm{Too \, optimisitic \, result} \nonumber
\end{equation}

\index{norm}{External information!priors} 
\index{norm}{External information!input errors} 
\index{norm}{External information!central values}
In many cases, the fit manifold is restricted by the knowledge 
from earlier experiments. For example, the knowledge on the solar 
parameters will in most cases be supplied by the solar neutrino experiments. 
If the external precision of a parameter is at the time of the 
measurement better than the one of the experiment itself, one usually will use
this external, better knowledge and impose a corresponding constraint
 on this parameter. This external knowledge may reduce the extension of the 
$n$-dimensional fit manifold in the respective direction. In the most 
extreme case, keeping all parameters but the measured one fixed 
in the analysis is equivalent to the assumption that all parameters 
are determined externally with infinitively high precisions. As this 
is a quite strong assumption, one should \emph{always} check the 
consequences of relaxing it and using realistic errors. Only if such
a test has demonstrated that the impact of the uncertainty on a 
given fit parameter is negligible, it can be assumed as fixed safely. 
 The inclusion of external input in \GLOBES\ is done by the use of 
Gau\ss ian {\em priors}: We assume that an external measurement has 
determined the measured parameter to be at the central value 
(which we call {\em central value}) with a $1 \sigma$ Gau\ss ian error 
(which we call {\em input error}). The explicit definition of these 
priors will be shown in the next section.

@@

\index{norm}{External information}
\index{norm}{External input| \see{External information}}
It is one of the strengths of the \GLOBES\ software to use external input 
in order to reduce the extension of the fit manifold with the knowledge 
from external
(earlier) measurements. The treatment of external input is done by the 
addition of Gau\ss ian so-called {\em priors} to the systematics-minimized 
$\chi^2$-function. For example, for the matter density, one obtains as the 
final projected $\chi^2_F$ after minimization over the matter density
scaling factor \index{norm}{Matter density!scaling factor} $\hat{\rho}$
\be
 \chi^2_F = \underset{\rho}{\mathrm{min}} \left( \chi^2(\hat{\rho}) +
 \frac{(\hat{\rho} - \hat{\rho}^0)^2}{\sigma_{\hat{\rho}}^2} \right).
\label{equ:priors}
\ee
This example is a very simple one, since in fact the
minimization is simultaneously performed over all priors and free oscillation 
parameters. In \equ{priors}, $\hat{\rho}^0$ is the {\em central value} of the 
prior, and $\sigma_{\hat{\rho}}$ the $1 \sigma$ absolute (half width) 
{\em input error}. Thus, it is assumed that an external measurement has 
determined the matter density with a precision (input error) 
$\sigma_{\hat{\rho}}$ at the central value $\hat{\rho}^0$. Usually, 
the central value is fixed at the best-fit value, and the input error 
to the $1 \sigma$ half width of the external measurement. For the matter 
density, $\hat{\rho}^0$  is usually set to $1.0$ corresponding to
the actual matter density profile such as given by the experiment
 definition file, and $\sigma_{\hat{\rho}}$ to the 
relative matter density uncertainty (\eg, $0.05$ for 5\% uncertainty).

In principle, one can set the priors for the matter density and all oscillation parameters. For example, if the disappearance channels of the experiment determine the leading oscillation parameters with unprecedented
precisions, one can omit the respective input errors. In \GLOBES , a
value of $0$ corresponds\footnote{To be precise, a value for the error
in between $-10^{-12}$ and $+10^{-12}$} to neglecting the prior. If, however,
 earlier external measurements provide better information, one can set their 
absolute precisions with the input errors. The central values are usually 
chosen to be the best-fit values of this external experiments, such as
for the input from solar experiments. In some cases, it may be
necessary to adjust them, such as for artificial constraints to the
oscillation parameters. 
For example, for the investigation of the
opposite-sign solution, one can use the prior to constrain $\ldm$
in order to force the minimizer not to  fall into the (unwanted) 
true-sign solution. In this case, the central value of $\ldm$ would 
be set to $\rho^0_{\ldm} = - \ldm$, and a $\sigma_{\ldm}$ of the order 
of $\ldm$ would be imposed. For the algorithm, it would 
then be rather difficult 
to converge into the unwanted true-sign solution. However, note that 
one should in this case check that the actually determined value 
for $\ldm$ after minimization is close enough to the guessed value $-\ldm$ 
in order to avoid significant artifical contributions of the priors to the 
final $\chi^2$. Alternatively one could re-run the minimizer with the
position of the previously found minimum as starting position but now
with switching off the constraint on $\ldm$.

\index{norm}{External information!input errors} 
\index{norm}{External information!central values}
In order to set the central values and input errors, two function have to
 be called {\em before the usage of any minimizer}:
\begin{function}
\GLBNS{SetCentralValues}
{\tt int glbSetCentralValues(const glb\_params in)} sets the central values for all of the following minimizer calls to {\tt in}.
\end{function}
\begin{function}
\GLBNS{SetInputErrors}
{\tt int glbSetInputErrors(const glb\_params in)}
 sets the input errors for all of the following minimizer calls to {\tt in}.
 An input error of $0$ corresponds to not taking into account the
 respective prior.
\end{function}
Accordingly, there are functions to return the actually set central values
and input errors:
\begin{function}
\GLBNS{GetCentralValues}
{\tt int glbGetCentralValues(glb\_params out)} writes the currently
set central values to {\tt out}.
\end{function}
\begin{function}
\GLBNS{GetInputErrors}
{\tt int glbGetInputErrors(glb\_params out)}
 writes the currently set input errors to {\tt out}.
\end{function}
All functions take or return as many matter density parameters as there are initialized experiments. In addition, they return $-1$ if the operation
was not successful.

\index{norm}{External information!precision}
Eventually, a typical initialization of the external input with
$10\%$ external precisions for the solar 
parameters\footnote{In fact, accelerator-based long-baseline experiments 
are primarily sensitive to the product $\sin 2 \theta_{12} \cdot \sdm$, 
which means that these errors effectively add up to an error of this 
product of about $15\%$.}, 
and $5\%$ matter density uncertainties for all experiments looks like this:
\begin{quote}
{\tt
 glbDefineParams(input\_errors,theta12*0.1,0,0,0,sdm*0.1,0);\\  
 glbSetDensityParams(input\_errors,0.05,GLB\_ALL);\\
 glbSetCentralValues(true\_values);\\
 glbSetInputErrors(input\_errors);\\
}
\end{quote}
In this example, the central values are set to 
the true (simulated) values. Remember that initially the 
matter density scaling factors are 
all $1.0$, which means that they do not need to be adjusted for the
central values.

\index{norm}{Minimizer!priors}
\index{norm}{External information!priors}
\index{norm}{Advanced tricks}
Though the priors are an elegant way to treat external input, 
there are also some complications with priors. The following hints are for the
more advanced \GLOBES\ user:
\begin{enumerate}
\item
 The priors are only added once to the final $\chi^2$, no matter how
 many experiments there are simulated. This is already one reason
 (besides the minimization) why the sum of all projected $\chi^2$'s of 
 the individual experiments 
 cannot correspond to the $\chi^2$ of the combination of all experiments.
\item
 Priors are not used for parameters which are not minimized over, \ie, 
 kept fixed. This will be important together with arbitrary projections
 using \GLB{ChiNP}. A more subtle consequence is the comparison
 of fit manifold sections and projections for the solutions where
 the absolute minimum $\chi^2$ is larger than zero, \ie, degeneracies
 other than the true solution. In this case, the sections and
 projections are not  comparable if not corrected by the prior contributions, where the
 correction can be obtained as the $\chi^2$-difference at the minimum.
 For example, projecting the $\mathrm{sgn}(\ldm)$-degeneracy onto
 the $\theta_{13}$-$\deltacp$-plane and comparing it with the section
 (all other parameters fixed), the section region would in many cases be 
 larger than the projection region if the priors were not added to the
 section. At the true solution, this problem usually does not occur
 because the prior contributions are close to zero.    
\item
Currently, \GLOBES\ only supports Gau\ss ian priors for the individual
oscillation parameters. Especially for the solar parameters, this is only
 an approximation, since they are imposed on $\theta_{12}$ and not
 on $\sin 2 \theta_{12}$, $\sin 2 \theta_{12} \cdot \sdm$, or $\sin  
 \theta_{12}$. \GLOBES\ 3.0 and higher provide an alternative
to that by the concept of user-defined priors.
 \end{enumerate}

@@

\begin{quote}
{\tt {\scriptsize
  /* Set central values and input errors for all projections */  \\
  glbDefineParams(input\_errors,theta12*0.1,0,0,0,sdm*0.1,0);\\  
  glbSetDensityParams(input\_errors,0.05,GLB\_ALL);\\
  glbSetCentralValues(true\_values);\\
  glbSetInputErrors(input\_errors);\\
\\
  /* Define my own two-parameter projection for glbChiNP: Only deltacp is free! */ \\
  glbDefineProjection(th13\_projection,GLB\_FIXED,GLB\_FIXED,GLB\_FIXED,GLB\_FREE,GLB\_FIXED,GLB\_FIXED); \\
  glbSetProjection(th13\_projection); \\ 
\\
  /* Iteration over all values to be computed */ \\
  double x,res1,res2;     \\
  for(x=-4;x<-2.0+0.001;x=x+2.0/50) \\
  \{ \\
\hspace*{0.5cm}  /* Set fit value of stheta */ \\
\hspace*{0.5cm} glbSetOscParams(test\_values,asin(sqrt(pow(10,x)))/2,1); \\
    \\
\hspace*{0.5cm} /* Guess fit value for deltacp in order to safely find minimum */ \\
\hspace*{0.5cm} glbSetOscParams(test\_values,200.0/2*(x+4)*M\_PI/180,3); \\
 \\
\hspace*{0.5cm} /* Compute Chi2 for user-defined two-parameter correlation */ \\
\hspace*{0.5cm} res1=glbChiNP(test\_values,NULL,GLB\_ALL); \\
      \\
\hspace*{0.5cm} /* Compute Chi2 for full correlation: minimize over all but theta13 */ \\
\hspace*{0.5cm} res2=glbChiTheta13(test\_values,NULL,GLB\_ALL); \\
  \\
\hspace*{0.5cm} AddToOutput(x,res1,res2);\\
  \} \\
